Some examples in extended logic counting classes

### 2016-10-05 QSO/cap-comp.tex
15 sep
5 sep

\begin{theorem}
	$\qfo(\tc)$ captures $\shl$ over the class of ordered structures.
\end{theorem}
\begin{proof}
	For the first condition, let $\alpha\in\qfo(\tc)$, let $\A\in\ostr$ $v$ and $V$ first and second order assignments of $\A$ respectively. We evaluate $\sem{alpha}(\A,v,V)$ recursively. A logic formula $\varphi$ is evaluated...?

@@

>>>>

\begin{theorem}
	$\qfo(\tc)$ captures $\shl$ over the class of ordered structures.
\end{theorem}
\begin{proof}
	For the first condition, let $\alpha\in\qfo(\tc)$, let $\A\in\ostr$ $v$ and $V$ first and second order assignments of $\A$ respectively. We evaluate $\sem{\alpha}(\A,v,V)$ recursively. A logic formula $\varphi$ is evaluated...?

@@

\begin{theorem}
	$\rqfo(\fo)$ captures $\fp$ over the class of ordered structures.
\end{theorem}

\begin{theorem}
	$\tqfo(\fo)$ captures $\shl$ over the class of ordered structures.
\end{theorem}
\begin{proof}
	For the second condition, let $f \in \shl$. Let $M$ be a non-deterministic logspace machine such that $f(w) = \acc_M(w)$ Let ${\cal Q} = \{q_1,\ldots,q_{\ell}\}$ be the set of states of $M$, where $q_1$ is the initial state, and let $q_{\ell}$ be the only final state of $M$. Let $\Sigma = \{0,1\}$ wlog. and let $w\in\Sigma^n$. Let $w = w_1\cdots w_n$, with $w_i \in \{0,1\}$ and let $\A$ be the ordered structure that represents this word with $A = \{1,\ldots,n\}$ and an unary predicate $P(\cdot)$ where $i\in P$ if and only if $w_i = 1$. We notate $M(w)$ as the graph of configurations of $M$ running on input $w$. Suppose that $M$ has a reading tape and a working tape, and the latter uses space $c\cdot\log(n)$.

	We represent configurations with a tuple of fixed size $k$. The formula $\varphi(\bar{x},\bar{y})$ describes consecutive configurations. The formula $\varphi_I(\bar{x})$ models that $\bar{x}$ is the initial configuration of $M(w)$. The formula $\varphi_F(\bar{x})$ models that $\bar{x}$ is a accepting (final) configuration of $M(w)$. The formula we construct is:
	$$
	\alpha = \sa{\bar{x}}\sa{\bar{y}}([\pth \varphi(\bar{u},\bar{v})](\bar{x},\bar{y})\cdot \varphi_I(\bar{x})\cdot\varphi_F(\bar{y})).
	$$
	Note that there is a total of $m = (n+2)\cdot (c\log(n)+2)\cdot 3^{c\log n} \cdot \ell$ distinct configurations in $M(w)$, so we will represent them in the following way. The first two elements of the tuple represent the position of the head in the reading tape. The next $e$ elements represent the position of the head in the working tape. We choose an appropriate $e$ such that $n^e \geq c\log(n)$ for each $n$. The next $2c$ elements represent the string in the working tape. The last $\ell$ elements represent the state in which the machine is in. To specify, let $\bar{a} = (a_1,\ldots,a_m)$ be a tuple that represents a configuration in $M(w)$.
	\begin{itemize}
	\item $(a_1,a_2)$ represent the position of the head in the reading tape. If $a_1 = 1$, then the head is in position $a_2$ (the $a_2$-th character in the input). If $a_1 = 2$, then $a_2 = 1$ represents that the head is at the beginning of the tape (position 0), and if $a_2 = 2$, then the head is at position $n+1$.
	\item $(a_3,\ldots,a_{e+2})$ represent the position of the head in the working tape. Specifically, let $h$ be the value of $a_{e+2}a_{e+1}\cdots a_4 a_3$ in base $n$. Then the head is at position $h$ of the working tape.
	\item $(a_{e+3},\ldots,a_{2+e+2c})$ represents the string that is in the working tape. Specifically, let $h$ be the value of $a_{2+e+2c}a_{1+e+2c}\cdots a_{e+4}a_{e+3}$ in base $n$. Then the string in the working tape is $h$ in base 3, where 2 is interpreted as $\texttt{B}$.
	\item $(a_{3+e+2c},\ldots,a_{2+e+2c+\ell})$ represents the current state $q_i$. Every value in the tuple is $1$ except the element $a_{3+e+2c+i}$ which is 2.
	\end{itemize}
	Now we describe the formulas.
	\begin{align*}
	\varphi_I(\bar{x}) &= \varphi_v(\bar{x}) \wedge \exists u(\varphi_1(u) \wedge \varphi_{succ}(u,x_{3+e+2c})) \\
	\varphi_F(\bar{x}) &= \varphi_v(\bar{x}) \wedge \exists u(\varphi_1(u) \wedge \varphi_{succ}(u,x_{2+e+2c+\ell})),
	\end{align*}
	where $\varphi_v(\bar{x})$ describes that $\bar{x}$ is a valid configuration of $M(w)$ and $\varphi_1(x)$ describes that $x$ is the first element of $A$.
\end{proof}

\begin{theorem}
	$\tqso(\fo)$ captures $\shp$ over the class of ordered structures.
\end{theorem}

\begin{theorem}
	($\qfo(\lfp)$ or $\qfo(\fo)$) with algebraic least fixed point operator captures $\totp$ over the class of ordered structures.
\end{theorem}

### 2016-10-05 QSO/logics.tex
5 sep

We define an operator which extends the Least Fixed Point logic to counting. Recall that a fixed point operator is defined by a formula $\varphi(x_1,\ldots,x_k,R)$ which is positive on $R$ where $R$ is a predicate of arity $k$. For a structure $\A$ with domain $A$, the operator $T_{\varphi}:2^{A^k} \to 2^{A^k}$ is defined as $T_{\varphi}(X) = \{(a_1,\ldots,a_k)\mid (\A,X)\models \varphi(a_1,\ldots,a_k,R) \}$, for each $X\subseteq A^k$. Let $T_0 = \emptyset$ and $T_{i+1} = T_{\varphi}(T_i)$ for each $i \in \nat$. Note that there exists $n\in \nat$ such that $T_{n+1} = T_n$. Semantically, $[\lfpop\,\varphi(x_1,\ldots,x_k,R)]$ is defined such that for each $(a_1,\ldots,a_k)\in A^k$, it holds that $\A\models[\lfpop\,\varphi(x_1,\ldots,x_k,R)](a_1,\ldots,a_k)$ if and only if $\A\models T_n(a_1,\ldots,a_k)$.

We extend $\qfo$ with the operator $[\textbf{alfp}\,\varphi(y_1,\ldots,y_k,R)\mid\alpha(x_1,\ldots,x_{\ell},R,\pi)](x_1,\ldots,x_{\ell})$. It is defined by a FO-formula $\varphi$ and a QFO-formula $\alpha$ which mentions a placeholder function $\pi$ with domain $\fv^{\ell}$. Let $\{T_i\}_{i\in\nat}$ and $n\in\nat$ be defined from $\varphi$ as it was for the $\lfp$ operator.

@@

To formally characterize the semantics of $[\textbf{alfp}\,\varphi(y_1,\ldots,y_k,R)\mid\alpha(x_1,\ldots,x_{\ell},R,\pi)](x_1,\ldots,x_{\ell})$. %. 
We name $\T = \{T_0,\ldots,T_n\}$ and we define a sequence of functions $\zeta_0,\ldots,\zeta_n:A^{\ell}\to\nat$ as follows. For each $T_i \in \T$, we define a $\qfo$ formula $\beta_i$ as follows. If $i = 0$, then $\beta_i(u_1,\ldots,u_{\ell}) = 0$. If $i \geq 1$, then
$$
\beta_i(u_1,\ldots,u_{\ell}) = \bigplus_{(a_1,\ldots,a_{\ell})\in A^{\ell}} \varphi_{a_1}(u_1)\cdot\varphi_{a_2}(u_2)\,\cdots\,\varphi_{a_{\ell}}(u_{\ell})\cdot \zeta_{i-1}(a_1,\ldots,a_{\ell}).
$$
Then, for each $T_i \in \T$, let $V$ be a second-order assignment for $\A$ that assigns $T_i$ to $R$, let $\zeta_i: A^{\ell}\to\nat$ be such that for each $(a_1,\ldots,a_{\ell})\in A^{\ell}$ it holds $\zeta_i(a_1,\ldots,a_{\ell}) = \sem{\alpha\mid_{\pi(u_1,\ldots,u_{\ell})\to \beta_i(u_1,\ldots,u_{\ell})}}(\A,v,V)$, where $v$ is a first-order assignment for $\A$ that satisfies $a_i = v(x_i)$ for each free $x_i$ in $\alpha$.

For a given first order asignment $v$ and a second order assignment $V$, let $a_i = v(x_i)$, the operator is then evaluated as:
$$
\sem{[\textbf{alfp}\,\varphi(y_1,\ldots,y_k,R)\mid\alpha(x_1,\ldots,x_{\ell},R,\pi)]}(\A,v,V) = \zeta_n(a_1,\ldots,a_{\ell}).
$$
As an example, we will show a formula that counts the number of paths of size $n$ of a structure $\A$ with a binary relation $E$. First we define $\varphi(x,R)$:
$$
\varphi(x,R) = \forall y(x < y \vee x = y) \vee \exists z(R(z) \wedge \varphi_{succ}(z,x)),
$$
where $\varphi_{succ}(x,y)$ is a formula that is satisfied by pairs $(x,y)$ that are consecutive in the order $<$. That is, $\varphi_{succ}(x,y) = x < y \wedge \forall z((x < z \wedge z < y) \to (x = z \vee z = y) )$. Now we define $\alpha(x,y,R,\pi)$ as:
$$
\alpha(x,y,R,\pi) = (\neg \exists zR(z))\cdot(x = y) + \Sigma z[\pi(x,z)\cdot E(z,y)].
$$
Then, the formula $[\alfp_{\varphi(x,R)} \alpha(x,y,R,\pi)](x,y)$ when evaluated on $(\A,a,b)$ will count the number of paths of size $n$ from $a$ to $b$.

We also define the less powerful operator $\pth$ as follows. Let $\A$ be an ordered structure. Given a formula $\psi(\bar{x},\bar{y})$, where $\vert \bar{x} \vert = \vert \bar{y} \vert = k$ let ${\cal G} = ({\cal V},\cal{E})$ be induced graph over the set of vertices ${\cal V} = A^k$, and for every $\bar{a},\bar{b}\in A^k$ it holds that ${\cal E}(\bar{a},\bar{b})$ if and only if $\A \models \psi(\bar{a},\bar{b})$. To formalize the semantics for this operator, let $n = \vert A^k \vert$.
For a given first order assignment $v$ and a second order asssignment $V$, let $\bar{a} = v(\bar{x})$ and $\bar{b} = v(\bar{y})$, and $\sem{[\pth\, \psi(\bar{x},\bar{y})]}(\A,v,V)$ will take the value of the number of paths of size $n$ from $\bar{a}$ to $\bar{b}$ in the graph ${\cal G}$.

\begin{theorem}
	Given a positive $\fo$ formula $\varphi(\bar{x},R)$ and a $\qfo$ formula $\alpha(\bar{x})$, there exists a $\qso$ formula $\beta(\bar{x})$ such that $\sem{[\alfp_{\varphi(\bar{x},R)} \alpha_{\psi}(\bar{x},R)](\bar{x})} = \sem{\beta(\bar{x})}$.
\end{theorem}
\begin{proof}

>>>>

We define an operator which extends the Least Fixed Point logic to counting. Recall that a fixed point operator is defined by a formula $\varphi(x_1,\ldots,x_k,R)$ which is positive on $R$ where $R$ is a predicate of arity $k$. For a structure $\A$ with domain $A$, the operator $T_{\varphi}:2^{A^k} \to 2^{A^k}$ is defined as $T_{\varphi}(X) = \{(a_1,\ldots,a_k)\mid (\A,X)\models \varphi(a_1,\ldots,a_k,R) \}$, for each $X\subseteq A^k$. Let $T_0 = \emptyset$ and $T_{i+1} = T_{\varphi}(T_i)$ for each $i \in \nat$. Note that there exists $n\in \nat$ such that $T_{n+1} = T_n$. Semantically, $[\lfpop\varphi(x_1,\ldots,x_k,R)]$ is defined such that for each $(a_1,\ldots,a_k)\in A^k$, it holds that $\A\models[\lfpop\varphi(x_1,\ldots,x_k,R)](a_1,\ldots,a_k)$ if and only if $\A\models T_n(a_1,\ldots,a_k)$.

We extend $\qfo$ with the operator $[\alfp\varphi(y_1,\ldots,y_k,R)\mid\alpha(x_1,\ldots,x_{\ell},R,\pi)](x_1,\ldots,x_{\ell})$. It is defined by a FO-formula $\varphi$ and a formula $\alpha$ in $\qfo$ without the $\Pi$ operator, which mentions a placeholder function $\pi$ with domain $\fv^{\ell}$. Let $\{T_i\}_{i\in\nat}$ and $n\in\nat$ be defined from $\varphi$ as it was for the $\lfp$ operator.

\martin{No estoy seguro si quitar el operador $\Pi$ de la fórmula o solo decir que la funcion $\pi$ no puede estar mencionada en ninguna instancia del operador $\Pi$.}

@@

To formally characterize the semantics of $[\alfp\varphi(y_1,\ldots,y_k,R)\mid\alpha(x_1,\ldots,x_{\ell},R,\pi)](x_1,\ldots,x_{\ell})$. %. 
We name $\T = \{T_0,\ldots,T_n\}$ and we define a sequence of functions $\zeta_0,\ldots,\zeta_n:A^{\ell}\to\nat$ as follows. For each $T_i \in \T$, we define a $\qfo$ formula $\beta_i$ as follows. If $i = 0$, then $\beta_i(u_1,\ldots,u_{\ell}) = 0$. If $i \geq 1$, then
$$
\beta_i(u_1,\ldots,u_{\ell}) = \bigplus_{(a_1,\ldots,a_{\ell})\in A^{\ell}} \varphi_{a_1}(u_1)\cdot\varphi_{a_2}(u_2)\,\cdots\,\varphi_{a_{\ell}}(u_{\ell})\cdot \zeta_{i-1}(a_1,\ldots,a_{\ell}).
$$
Then, for each $T_i \in \T$, let $V$ be a second-order assignment for $\A$ that assigns $T_i$ to $R$, let $\zeta_i: A^{\ell}\to\nat$ be such that for each $(a_1,\ldots,a_{\ell})\in A^{\ell}$ it holds $\zeta_i(a_1,\ldots,a_{\ell}) = \sem{\alpha\mid_{\pi(u_1,\ldots,u_{\ell})\to \beta_i(u_1,\ldots,u_{\ell})}}(\A,v,V)$, where $v$ is a first-order assignment for $\A$ that satisfies $a_i = v(x_i)$ for each free $x_i$ in $\alpha$.

For a given first order asignment $v$ and a second order assignment $V$, let $a_i = v(x_i)$, the operator is then evaluated as:
$$
\sem{[\alfp\varphi(y_1,\ldots,y_k,R)\mid\alpha(x_1,\ldots,x_{\ell},R,\pi)]}(\A,v,V) = \zeta_n(a_1,\ldots,a_{\ell}).
$$
As an example, we will show a formula that counts the number of paths of size $n$ of a structure $\A$ with a binary relation $E$. First we define $\varphi(x,R)$:
$$
\varphi(x,R) = \forall y(x < y \vee x = y) \vee \exists z(R(z) \wedge \varphi_{succ}(z,x)),
$$
where $\varphi_{succ}(x,y)$ is a formula that is satisfied by pairs $(x,y)$ that are consecutive in the order $<$. That is, $\varphi_{succ}(x,y) = x < y \wedge \forall z((x < z \wedge z < y) \to (x = z \vee z = y) )$. Now we define $\alpha(x,y,R,\pi)$ as:
$$
\alpha(x,y,R,\pi) = (\neg \exists zR(z))\cdot(x = y) + \Sigma z[\pi(x,z)\cdot E(z,y)].
$$
Then, the formula $[\alfp\varphi(x,R)\mid \alpha(x,y,R,\pi)](x,y)$ when evaluated on $(\A,a,b)$ will count the number of paths of size $n$ from $a$ to $b$.
This operator lets us define the set of recursive $\qfo$ formulas over $\R$ ($\rqfo$-formulas) using the following grammar:
\begin{eqnarray*}
\label{eq-def-rqfo}
\alpha & := & \varphi \ \mid \ s \ \mid \ (\alpha \add \alpha) \ \mid\ (\alpha \mult \alpha) \ \mid \ \sa{x} \alpha \ \mid \ \pa{x} \alpha \ \mid \ [\alfp \varphi \mid \alpha]
\end{eqnarray*}
where $\varphi$ is an $\fo$-formula over $\R$, $s \in \bbN$ and $x \in \fv$.

We also define the less powerful operator {\bf path} as follows. Let $\A$ be an ordered structure. Given a formula $\psi(\bar{x},\bar{y})$, where $\vert \bar{x} \vert = \vert \bar{y} \vert = k$ let ${\cal G} = ({\cal V},\cal{E})$ be induced graph over the set of vertices ${\cal V} = A^k$, and for every $\bar{a},\bar{b}\in A^k$ it holds that ${\cal E}(\bar{a},\bar{b})$ if and only if $\A \models \psi(\bar{a},\bar{b})$. To formalize the semantics for this operator, let $n = \vert A^k \vert$.
For a given first order assignment $v$ and a second order asssignment $V$, let $\bar{a} = v(\bar{x})$ and $\bar{b} = v(\bar{y})$, and $\sem{[\pth\, \psi(\bar{x},\bar{y})]}(\A,v,V)$ will take the value of the number of paths of size less or equal to $n$ from $\bar{a}$ to $\bar{b}$ in the graph ${\cal G}$. This operator lets us define the set of transitive $\qfo$ formulas over $\R$ ($\tqfo$-formulas) using the following grammar:
\begin{eqnarray*}
	\label{eq-def-tqfo}	\alpha & := & \varphi \ \mid \ s \ \mid \ (\alpha \add \alpha) \ \mid\ (\alpha \mult \alpha) \ \mid \ \sa{x} \alpha \ \mid \ \pa{x} \alpha \ \mid \ [\pth \varphi]
\end{eqnarray*}
where $\varphi$ is an $\fo$-formula over $\R$, $s \in \bbN$ and $x \in \fv$.

We also define the set of transitive $\qso$ formulas ($\tqso$-formulas) using the following grammar:
\begin{eqnarray*}
	\label{eq-def-tqso}	\alpha & := & \varphi \ \mid \ s \ \mid \ (\alpha \add \alpha) \ \mid\ (\alpha \mult \alpha) \ \mid \ \sa{x} \alpha \ \mid \ \pa{x} \alpha \ \mid \ \sa{X} \alpha \ \mid \ \pa{X} \alpha \ \mid \ [\pth \varphi]
\end{eqnarray*}
where $\varphi$ is an $\so$-formula over $\R$, $s \in \bbN$, $x \in \fv$ and $X \in \sv$.
\begin{theorem}
	Given a positive $\fo$ formula $\varphi(\bar{x},R)$ and a $\qfo$ formula $\alpha(\bar{x})$, there exists a $\qso$ formula $\beta(\bar{x})$ such that $\sem{[\alfp\varphi(\bar{x},R)\mid \alpha(\bar{x},R)](\bar{x})} = \sem{\beta(\bar{x})}$.
\end{theorem}
\begin{proof}
